---
title:  "Data Science Project using R Studio"
author: "Tochi Agbara, Karen Harker, Kevin Brown"
editor: visual
---

# Business Problem

We will work with The Violence Project's Mass Shooter Database and the purpose of this project is to better understand the underlying factors, patterns, and consequences associated with public mass shootings, which could aid in the development of preventive strategies, interventions, and support for survivors.  The data will be examined at three levels: the perpetrator, the victims, and the locations or communities.  The desired outcomes will be models of typologies of each of these levels that could be used to generate new research questions, risk assessment models, prevention interventions, and trauma interventions for survivors and witnesses.   

**Methods**

We would typically use machine learning algorithms, such as classification models, or bootstrapping to build more data resample with replacement. These will help to build a predictive model for identifying individuals at risk of perpetrating shooting incidents. We could also do a correlation test of mass shooters in Texas to ascertain the gun control level. We are focusing on using unsupervised learning by clustering analysis using age, criminal background, ethnicity, education, etc. to help the policymakers to minimize casualties. Finally, suggestions for future research on the relationship between homicide and the Internet will be considered (Patricia, 2021). 

## Read the data / Import the Data

We import the Full Database sheet and the historical state population by year then converted to csv...

```{r, echo=TRUE}
library(readr)
library(dplyr)
library(ggplot2)
library(corrplot)
library(lubridate)
library(tidyr)
library(GGally)

Full_Database<-read.csv("Full_Database.csv", header=TRUE)
historical_state_population_by_year<-read.csv("historical_state_population_by_year.csv", header=TRUE)

# KH: read the State Firearms Law table that provides sums of restrictive gun laws for that state in the year of that event.
sfl_sum <- read.csv("SFL_Sum_Year_by_MSDB_Case.csv", header=TRUE)

names(Full_Database)
names(historical_state_population_by_year)
names(sfl_sum)
```

## Join / Merge the data

We want to join the data on state and year to gain insights intto state population in a given time period

```{r, echo=TRUE}

# Merge data and keep all rows from Full_Database
merged_data <- Full_Database %>%
    left_join(historical_state_population_by_year, by=c("State" = "PopulationState", "Year" = "StatePopulationYear"))

# KH: joined with the SFL dataset, adding SLF_Sum_Year column
merged_data <- merged_data %>%
   left_join(sfl_sum, by="Case..")

names(merged_data)
merged_data
```

## Clean the Data

```{r}

```

## EDA

```{r}

```

## Statistical Analysis

```{r}

```

## Data Modeling

```{r}

```

## Cluster Analysis

```{r}

```

## Boot strapping

```{r}

```

## Data Reporting & Communication Results

## 
